Inter-rater reliability of the tracking and reporting outcomes of procedural sedation tool

### 2021-05-25 manuscript.Rmd
submitted to jrn

Clinical guidelines recommend that outcomes of procedural sedation are audited. The Tracking and Reporting Outcomes of Procedural Sedation (TROOPS) tool, developed by the International Consortium for the Advancement of Procedural Sedation, is intended to be suitable for use in all locations by sedation providers. We evaluated the feasibility and reliability of using the TROOPS tool for sedation that is administered by nurses in the cardiac catheterization laboratory. The version of TROOPS we used contained items in the minor, intermediate and sentinel categories. Two nurses independently completed the TROOPS tool for 40 patients who underwent a procedure with sedation. A sedation-related adverse event was selected by at least one of the raters for 21 of the patients (52%; 95% CI = 0.38 to 0.67). Most were minor events related to the airway and breathing category of the TROOPS tool (n = 17; 42%; 95% CI = 0.29 to 0.58). The remaining were intermediate-severity events related to sedation quality. No events for the other categories of the TROOPS tool were selected. The intra-class correlation coefficient between paired nurse ratings of the TROOPS tool was `r total` (95% CI = `r total_lower` to `r total_upper`). Most nurses (85%, 95% CI = 73% to 92%) reported that it took less than 1 minute to complete the tool. In summary, paired ratings of sedation-related adverse events by nurses achieved moderate consistency using a version of the TROOPS tool containing items from the minor category of the airway and breathing domain. Use of this tool in clinical practice seems feasible.

@@

A data collection form that contained three parts was developed to be used in this study. The first part contained items to collect demographic (age, sex) and clinical information (type of procedure, procedure start and end time, baseline oxygen supplementation and total doses of sedation used) about the included patients.

The second part consisted of the quality improvement version of the TROOPS tool, with the addition of minor adverse events from the 'airway and breathing' as well as 'gastrointestinal' domains from the research version of the tool.[@roback2018tracking] Minor events from the 'airway and breathing' domain used were 'increased or added supplemental oxygen', 'airway repositioning' and 'tactile stimulation'. Based on consultation with members of the clinical team, we elected not to include the minor items related to hypersalivation (suctioning for hypersalivation and anticholinergic for hypersalivation) because the medications used for sedation in the department (fentanyl and midazolam) are not associated with this particular adverse effect. Both minor events from the gastrointestinal domain of the research version were used. These were administration of anti-emetic and suction for emesis. The minor items from the allergy and neurological domains were not included because these adverse effects were not considered to occur frequently enough in this particular sedation context to warrant inclusion as a measure for quality improvement. Other modifications were made to the TROOPS tool through consultation with members of the clinical team in order to make it suitable for use by nurses in that particular department. Modifications included the removal of outcomes that were not feasibly assessed by nurses in the immediate post-procedure period, including permanent neurological deficit from the sentinel category of the neurological domain and pulmonary aspiration syndrome from the sentinel adverse events category of the airway and breathing domain.

@@

Inter-rater reliability was assessed by calculating the random effects one-way intra-class correlation between nurses’ paired ratings. Lower and upper confidence(95%) bounds associated with the intra-class correlation under the one-factor ANOVA model was used because each patient could have been rated by a different group of raters (i.e. nurses).[@gwet2014handbook] Feasibility was examined by calculating proportions for nurses' selections about the time taken to complete the tool. Analyses were undertaken using `R`.[@rcore2021] Data, statistical code and a computational environment required to reproduce the results is available.[insert ref]

@@

Data were collected from January to November, 2017. Two nurses independently completed the TROOPS tool for 40 patients who underwent a procedure with sedation. A summary of the patient characteristics is presented in Table \@ref(tab:tab1). Most patients were undergoing a cardiac implantable electronic device implant (e.g. pacemaker or implantable cardioverter defibrillator) or electrophysiology procedure. A combination of midazolam and fentanyl was used for all patients.

@@

In our study, both nurses who completed TROOPS ratings for each patient were involved in the procedure, but they performed different roles. In the department in which data collection was performed, nurses performed the roles of 'scout' nurse and 'monitor' nurse. This is commonplace in the cardiac catheterization laboratory setting.[@white2018consensus] Based on the authors' experiences of working in cardiac catheterization laboratory settings, often the responsibility for documentation, including that related to audits for quality improvement activities, is sometimes shared between the nurses performing these roles in an very inconsistent manner. Our findings show that TROOPS ratings for the same procedure were not in perfect agreement between nurses who performed these different roles. One course of action that would logically be expected to improve the reliability of measuring adverse sedation events using the TROOPS tool is to ensure that the rater is consistently only ever the nurse who performed the scout role *or* the nurse who performed the monitor role.

>>>>

Clinical guidelines recommend that procedural sedation outcomes should be audited. The Tracking and Reporting Outcomes of Procedural Sedation (TROOPS) tool, developed by the International Consortium for the Advancement of Procedural Sedation, is intended to be suitable for use in all locations by sedation providers. We evaluated the feasibility and reliability of using the TROOPS tool for sedation that is administered by nurses in the cardiac catheterization laboratory. The version of TROOPS we used contained items in the minor, intermediate and sentinel categories. Two nurses independently completed the TROOPS tool for 40 patients who underwent a procedure with sedation. A sedation-related adverse event was selected by at least one of the raters for 21 of the patients (52%; 95% CI = 0.38 to 0.67). Most were minor events related to the airway and breathing category of the TROOPS tool (n = 17; 42%; 95% CI = 0.29 to 0.58). The remaining were intermediate-severity events related to sedation quality. No events for the other categories of the TROOPS tool were selected. The intra-class correlation coefficient between paired nurse ratings of the TROOPS tool was `r total` (95% CI = `r total_lower` to `r total_upper`). Most nurses (85%, 95% CI = 73% to 92%) reported that it took less than 1 minute to complete the tool. In summary, paired ratings of sedation-related adverse events by nurses achieved moderate consistency using a version of the TROOPS tool containing items from the minor category of the airway and breathing domain. Use of this tool in clinical practice seems feasible.

@@

A data collection form that contained three parts was developed to be used in this study. The first part contained items to collect demographic (e.g., age, sex) and clinical information (i.e., type of procedure, procedure start and end time, baseline oxygen supplementation and total doses of sedation used) about the included patients.

The second part consisted of the quality improvement version of the TROOPS tool, with the addition of minor adverse events from the 'airway and breathing' as well as 'gastrointestinal' domains from the research version of the tool.[@roback2018tracking] Minor events from the 'airway and breathing' domain used were 'increased or added supplemental oxygen', 'airway repositioning' and 'tactile stimulation'. Based on consultation with members of the clinical team, we elected not to include the minor items related to hypersalivation (i.e., suctioning for hypersalivation and anticholinergic for hypersalivation) because the medications used for sedation in the department (i.e., fentanyl and midazolam) are not associated with this particular adverse effect. Both minor events from the gastrointestinal domain of the research version were used. These were administration of anti-emetic and suction for emesis. The minor items from the allergy and neurological domains were not included because these adverse effects were not considered to occur frequently enough in this particular sedation context to warrant inclusion as a measure for quality improvement. Other modifications were made to the TROOPS tool through consultation with members of the clinical team in order to make it suitable for use by nurses in that particular department. Modifications included the removal of outcomes that were not feasibly assessed by nurses in the immediate post-procedure period, including permanent neurological deficit from the sentinel category of the neurological domain and pulmonary aspiration syndrome from the sentinel adverse events category of the airway and breathing domain.

@@

Inter-rater reliability was assessed by calculating the random effects one-way intra-class correlation between nurses’ paired ratings. Lower and upper confidence (95%) bounds associated with the intra-class correlation under the one-factor ANOVA model was used because each patient could have been rated by a different group of raters (i.e. nurses).[@gwet2014handbook] Feasibility was examined by calculating proportions for nurses' selections about the time taken to complete the tool. Analyses were undertaken using `R`.[@rcore2021] Data, statistical code and a computational environment required to reproduce the results is available.[@conway2021code]

@@

Data were collected from January to November, 2017. Two nurses independently completed the TROOPS tool for 40 patients who underwent a procedure with sedation. A summary of the patient characteristics is presented in Table \@ref(tab:tab1). Most patients were undergoing a cardiac implantable electronic device implant (e.g. pacemaker or implantable cardioverter defibrillator) or electrophysiology procedure (e.g., atrial fibrillation or ventricular tachycardia ablation). A combination of midazolam and fentanyl was used for all patients.

@@

In our study, both nurses who completed TROOPS ratings for each patient were involved in the procedure, but they performed different roles. In the department in which data collection was performed, nurses performed the roles of 'scout' nurse and 'monitor' nurse. This is commonplace in the cardiac catheterization laboratory setting.[@white2018consensus] Based on the authors' experiences of working in cardiac catheterization laboratory settings, often the responsibility for documentation, including that related to audits for quality improvement activities, is sometimes shared between the nurses performing these roles in a very inconsistent manner. Our findings show that TROOPS ratings for the same procedure were not in perfect agreement between nurses who performed these different roles. One course of action that would logically be expected to improve the reliability of measuring adverse sedation events using the TROOPS tool is to ensure that the rater is consistently only ever the nurse who performed the scout role *or* the nurse who performed the monitor role.
